---
title: "Data Cleaning"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages needed here, warning= FALSE, message = FALSE}
#be careful about suppressing warning messages. 
library(dplyr)
library(tidyr)
```

you can call a function explicitly from a package with package_name::funtion_name()
^do this if you the package isn't doing what you expect i.e. there are two packages with the same named function

# Read in the data

```{r}
catch_original <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", stringsAsFactors = FALSE)

#windows computers may need this code: read.csv(url("link", method = "libcurl"), stringsAsFactors = FALSE)

head(catch_original)

#command + shift + m for pipe operator %>% 





```


# clean data - make tidy

Remove notes and "all" columns becasue we don't need them
"I" in the chinook column (row 401) should be a 1

```{r}
catch_clean <- catch_original %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum) %>% 
  # to fix chinook column
  mutate(Chinook = ifelse(Chinook =="I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook))

# you could also select by clumn you want to delete with select(-All, -notesRegCode) this drops just the notesRegCode column

head(catch_clean)
summary(catch_clean)
# summary lets you see that chinook column is a character

```



# data QA

Which values were converted to NA by as.numeric
Checking for NA calues

```{r}
#which values in the Chinook column that are problematic ?
i <- which(is.na(catch_clean$Chinook))
i
catch_original[i, ]
#This shows that there was an I in the Chinook column on row 401 
```


## convert to long format

```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = c(Chinook, Sockeye, Coho, Pink, Chum), 
               names_to = "species", values_to = "catch") %>% 
  rename(catch_thousands = catch) %>% 
  mutate(catch = catch_thousands * 1000) %>% 
  #to drop the column after tranforming catch to thousands metric
  select(-catch_thousands)
  #could use -c(Region, Year) instead of list if you have a bunch 

head(catch_long)
```

# Example of pivot_wider


Eaxmple for wide format
```{r}
catch_wide <- catch_long %>% pivot_wider(names_from = "Year"
                                         , values_from = "catch")

head(catch_wide)
```


# summarize

mean catch per region

```{r}
mean_region <- catch_long %>%
  group_by(Region, species) %>% 
  summarise(mean_catch = mean(catch),
            n_obs = n())

head(mean_region)
  
```

FIlter for only sockeye species
```{r}
```


```{r}
sockeye_catch <- catch_long %>% 
  filter(species =="Sockeye") %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) 
  

head(sockeye_catch)
```


# join to other data

Read in the region definitions table
```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1", stringsAsFactors = FALSE) %>% select(code, mgmtArea)
head(region_defs)


```

```{r}
catch_joined <- left_join(catch_long, region_defs,
                          by = c("Region" = "code"))
head(catch_joined)
```


# seperate and unite

```{r}
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAI-039"),
                       stringsAsFactors = FALSE)


sites_df %>% 
  separate(site, into = c("island", "site_number"), sep = "-")

# can set remove = false to keep original column
```


```{r}
#Unite example
dates_df <- data.frame(year = c("1930",
                                "1930",
                                "1930"),
                       month = c("12",
                                "12",
                                "12"),
                       day = c("14",
                               "15",
                               "16"),
                       stringsAsFactors = FALSE)

dates_df %>% 
  unite(date, year, month, day, sep = "-", remove = FALSE)
```





